Combining predicate transformer semantics for effects

### 2020-08-29 msfp2020/msfp2020.tex
tweaks

\pause
The generic fold that computes a predicate of type \Set:
\begin{minted}{Agda}
[[_]] : Free C R a -> ((c : C) -> (R c -> Set))
      -> (a -> Set) -> Set
[[ Pure x ]] alg P = P x
[[ Op c k ]] alg P = alg c (λ x -> [[ k x ]] alg P)
\end{minted}
\end{frame}

@@

We implement \match as a case distinction.
\begin{minted}{Agda}
match : (r : Regex) -> String -> Nondet (Tree r)
match Empty         xs       = Op Fail λ()
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match Epsilon       Nil      = Pure tt
match Epsilon       (_ :: _) = Op Fail λ()
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (Singleton c) xs       =
    if xs = [c] then Pure c else Op Fail λ()
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (l | r)       xs       = Op Choice (λ b ->
    if b then Inl <$> match l xs else Inr <$> match r xs)
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (l · r) xs = do
    (ys, zs) <- allSplits xs
    (,) <$> match l ys <*> match r zs
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (r *) xs = match (r · (r *)) xs
\end{minted}
\pause
\alert{Error: \Agda{match (r *) xs} does not terminate}

@@

The missing ingredient is the rule of consequence:
\begin{minted}{Agda}
consequence : ∀ pt (S : Free es a) (f : a → Free es b) →
    [[ S ]] pt (λ x → [[ f x ]] pt P) ≡ [[ mx >>= f ]] pt P
\end{minted}

@@

Now we can finish the definition and prove soundness unconditionally:
\begin{minted}{Agda}
match (r *) = Op iRec (r · (r *))

@@

\begin{frame}[fragile]{Defining a derivative-based matcher}
To guarantee termination, use recursion on \Agda{xs} rather than \Agda{r}.\\
The \emph{Brzozowski derivative} \Agda{d r /d x} matches \Agda{xs} iff \Agda{r} matches \Agda{x :: xs};\\ \Agda{integralTree r : tree (d r /d x) -> tree r} ``integrates'' parse trees.

\begin{minted}{Agda}
dmatch : (r : Regex) -> Free es (tree r)
dmatch r = symbol >>= maybe
    (λ x -> Op iRec (d r /d x) (integralTree r))
    (if p <- matchEpsilon r
     then Pure (Sigma.fst p)
     else Op iND Fail λ())
\end{minted}

@@

\begin{minted}{Agda}
terminates-in : (f : (i : I) -> Free (Rec I O :: es) (O i))
    (S : Free (Rec I O :: es) a) → ℕ → Set
terminates-in f (Pure x)           n        = ⊤
terminates-in f (Op ∈Head c k)     Zero     = ⊥
terminates-in f (Op ∈Head c k)     (Succ n) =
    terminates-in pt f (f c >>= k) n
terminates-in f (Op (∈Tail i) c k) n        =
    pts i c (λ x → terminates-in f (k x) n)
\end{minted}
\end{frame}

>>>>

\pause
The generic fold that computes a predicate of type \Set:
\begin{minted}{Agda}
[[_]] : Free C R a -> ((c : C) -> (R c -> Set))
      -> (a -> Set) -> Set
[[ Pure x ]] alg P = P x
[[ Op c k ]] alg P = alg c (λ x -> [[ k x ]] alg P)
\end{minted}

\pause
Let us look at \Agda{alg : (c : C) -> (R c -> Set) -> Set}'s type.
\end{frame}

@@

We implement \match as a case distinction.
\begin{minted}{Agda}
match : (r : Regex) -> String -> Nondet (Tree r)
match Empty         xs       = Op Fail λ()
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match Epsilon       Nil      = Pure tt
match Epsilon       (_ :: _) = Op Fail λ()
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (Singleton c) xs       =
    if xs = [c] then Pure c else Op Fail λ()
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (l | r)       xs       = Op Choice (λ b ->
    if b then Inl <$> match l xs else Inr <$> match r xs)
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (l · r) xs = do
    (ys, zs) <- allSplits xs
    (,) <$> match l ys <*> match r zs
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (r *) xs = match (Epsilon | r · (r *)) xs
\end{minted}
\pause
\alert{Error: \Agda{match (r *) xs} does not terminate}

@@

The missing ingredient is the rule of consequence:
\begin{minted}{Agda}
consequence : ∀ pt (S : Free es a) (f : a -> Free es b) ->
    [[ S >>= f ]] pt P ≡ [[ S ]] pt (λ x -> [[ f x ]] pt P)
\end{minted}

@@

Now we can finish the definition and prove soundness unconditionally:
\begin{minted}{Agda}
match (r *) = Op iRec (Epsilon | r · (r *)) eitherToList

@@

\begin{frame}[fragile]{Defining a derivative-based matcher}
To guarantee termination, use recursion on \Agda{xs} rather than \Agda{r}.\\
The \emph{Brzozowski derivative} \Agda{d r /d x} matches \Agda{xs} iff \Agda{r} matches \Agda{x :: xs}.

\begin{minted}{Agda}
dmatch : (r : Regex) -> Free es (Tree r)
dmatch r = Op iParse
    λ { (Just x) -> Op iRec (d r /d x) (integralTree r)
        Nothing  -> if p <- matchEpsilon r
                    then Pure (Sigma.fst p)
                    else Op iND Fail λ() }
\end{minted}

\Agda{integralTree r : tree (d r /d x) -> tree r} ``integrates'' parse trees.

@@

\begin{minted}{Agda}
terminates-in : (f : (i : I) -> Free (Rec I O :: es) (O i))
    (S : Free (Rec I O :: es) a) → ℕ → Set
terminates-in f (Pure x)           n        = ⊤
terminates-in f (Op ∈Head c k)     Zero     = ⊥
terminates-in f (Op ∈Head c k)     (Succ n) =
    terminates-in pt f (f c >>= k) n
terminates-in f (Op (∈Tail i) c k) n        =
    pts i c (λ x -> terminates-in f (k x) n)
\end{minted}
\end{frame}
